---
title: "6.3. Analysing proportions"
subtitle: "The binomial test"
author: "Bárbara D. Bitarello"
date: "11-18-2025"
keywords: "Binomial test"
format:
  revealjs:
    hash-type: number
    embed-resources: false
    smaller: true
    scrollable: false
    incremental: true
    slide-number: true
    transition: slide
    background-transition: fade
    highlight-style: ayu-dark
    footer: "B215: Biostatistics with R"
    center-title-slide: false
    theme: [custom.scss]
    callout-appearance: simple
    code-fold: false
    chalkboard: true
    auto-slide: 133000
code-link: true
execute:
  echo: false
  freeze: auto
knitr: 
  opts_chunk:
    tidy: true
    tidy-opts: { "value": "list(width.cutoff = I(60))", "tag": "!expr" }
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=T)
library(tidyverse)
library(forcats)
library(dplyr)
library(ggplot2)
library(ggforce)
library(tidyr)
library(stringr)
library(knitr)
library(ggthemes)
library(wesanderson)
library(gt)
library(RColorBrewer)
library(plotly)
library(readr)
library(thematic)
library(infer)
library(kableExtra)
library(gridExtra)
library(emoGG)
library(emo)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
source("bb_theme.R")

# set the theme
theme_set(bb_theme())
```

## Binomial Test: Recap {visibility="hidden"}

For the following scenarios, state whether the binomial distribution would describe the probability distribution of possible outcomes. If not, explain why not.

    The number of red cards out of the first five cards drawn from the top of a regular, randomly shuffled deck of cards.
    The number of red balls out of 10 drawn one by one from a vat of 50 red and blue balls, if the balls are replaced and mixed after each draw.
    The number of red balls out of 10 drawn one by one from a vat of 50 red and blue balls, if the balls are not replaced after each draw.
    The number of red-eyed flies among 200 Drosophila individuals drawn at random from a large population having both red- and black-eyed flies.
    The total number of red-eyed flies in five Drosophila families, each of 40 individuals, with the families chosen at random from a large population.

## Properties of the binomial distribution {.dark-theme .center}

## Expected \# successes of n trials

Expected successes in $n$ trials

$$\Huge{\mu = n \times p}$$

$\mu$: That's right. Mean number of successes, aka, expected number of
successes.

::: aside
Probability Trivia: A binomial with n = 1 is called a Bernoulli process.
This will not be a factoid you need to know in this class, but you might
hear it some time.
:::

## Binomial: Variability and Uncertainty

For counts:

$$\text{population variance: } \sigma^2 = n \times p \times(1-p)$$

$$\text{sample variance: }s^2 = n\times \hat p \times(1-\hat p)$$

::: aside
Note: Some textbooks use n-1 rather than n for the denominator of the standard error. Dividing by n-1 decreases bias, but yields estimates results  that are further from the truth, on average.
:::

## Properties of proportions: expectations

If there are $X$ successes in $n$ trials in a random sample, then the

*estimated proportion of successes* is:

$$\hat p = \frac{X}{n}$$

$\hat p$: The hat signals that this is an estimate

$p$: The expected proportion of successes is the probability of success p



## Sample size and standard error

```{r sampdist}
set.seed(1234)
p10<-tibble(x=rbinom(size=10, p=0.2, n=1e6)/10) |> ggplot(aes(x=x)) + geom_histogram(bins =10, color="white") + #scale_x_continuous(breaks = 0:10)+
  ggtitle("sampling distribution", subtitle="(n=10 trials; p=0.2)")+xlab("Sample proportion")+
  annotate(geom = "text", x=0.5, y=1e5, label="SE=0.13\n s of samp.dist=0.13")+
  theme_minimal() 
p300<-tibble(x=rbinom(size=300, p=0.2, n=1e6)/300) |> 
  ggplot(aes(x=x)) + 
  geom_histogram(bins =50, color="white")+ 
  ggtitle("sampling distribution", subtitle="(n=300 trials; p=0.2)")+
  xlab("Sample proportion")+
  annotate(geom = "text", x=0.25, y=1e5, label="SE=0.023\ns of samp. dist = 0.023")+
  theme_minimal() 
library(patchwork)
p10+p300
```

## Properties of Proportions: Variability

|                           |                                          |     |
|---------------------------|------------------------------------------|-----|
| Variance                  | $\sigma^2=p(1-p)$                        |     |
| Standard deviation        | $\sigma=\sqrt(p(1-p))$                   |     |
| Sample variance           | $s^2=\hat p(1-\hat p)$      |     |
| Sample standard deviation | $s=\sqrt{\hat p(1-\hat p)}$ |     |

## Standard error of a proportion

Recall: the standard error of an estimate is always the standard
deviation of its sampling distribution. If we know the real $p$ we can
calculate it as:

$$\sigma_{\hat p}=\sqrt{\frac{p(1-p)}{n}}$$

Because $p$ is usually unknown, in practice we have:

$$s_{\hat p}=\sqrt{\frac{\hat p(1-\hat p)}{n}}$$

::: aside
Note: Some textbooks use $n-1$ rather than n for the denominator of the standard error. Dividing by $n-1$ decreases bias, but yields estimates results  that are further from the truth, on average.
:::

## 95% CIs for a proportion (1/)

::::: nonincremental
-   For many parameter values, the binomial distribution is not
    symmetric about its mean. This is one reason why the 2 SE rule of
    thumb does not work for this.
-   The 95% confidence interval of a proportion will enclose the true
    value of the proportion 95% of the time that it is calculated from
    new data.
-   There are many ways to *approximate* the 95% CI of a proportion, and
    no consensus\[1\].
:::

::: aside
\[1\]: In `R` "*Confidence intervals are obtained by a procedure first
given in Clopper and Pearson (1934). This guarantees that the confidence
level is at least conf.level, but in general does not give the
shortest-length confidence intervals.*" (from `help(binom.test)` in `R`)
:::

## 95% CIs for a proportion (2/)

-   The book suggests the Agresti-Coull CIs:

. . .

$$\bigg(p' - 1.96 \sqrt{\frac{p'(1-p')}{n+4}}\bigg) \leq p \leq \bigg(p' + 1.96 \sqrt{\frac{p'(1-p')}{n+4}}\bigg)$$

 $p' = (X+2) / (n+4)$...
    where $X  =$ observed \# of successes
    

::: aside
\[1\]: In `R` "*Confidence intervals are obtained by a procedure first
given in Clopper and Pearson (1934). This guarantees that the confidence
level is at least conf.level, but in general does not give the
shortest-length confidence intervals.*" (from `help(binom.test)` in `R`)
:::



## Approximations for the Binomial Test {.dark-theme .center}


## Why?

• Adding up all the probabilities to obtain the exact p-
value is pretty time-consuming

• Because of that, two approximations for the p-value of
a binomial test exist: the chi-square goodness-of-fit
test (next topic) and the normal approximation to the
binomial test

• They can be useful and are very common.

• In R you can implement the exact binomial test easily,
though, as we saw

::: aside
If asked to conduct a binomial test by hand, you can use the chi-squared approximation to get a p-value
:::

## Key Points

• A proportion is an estimate of a probability.

• The binomial distribution provides a clear
example of building up from simple probability
logic to mathematical models of probability.

• The binomial test tests the null hypothesis that
an estimated proportion comes from a
population with a specified probability.

• We can quantify the uncertainty in our estimate
of the probability.

Interactive App: https://shiny.rit.albany.edu/stat/binomial/

