---
title: "Examples from lectures (ch. 8-9)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=T, tidy="styler")
library(tidyverse)
library(knitr)
library(ggthemes)
library(wesanderson)
library(gt)
library(RColorBrewer)
library(plotly)
library(thematic)
library(infer)
library(kableExtra)
library(gridExtra)

#source("bb_theme.R")
options(digits = 3)


#theme_set(bb_theme())
theme_set(theme_minimal())


# color-blind friendly color palette
okabe_ito<-c(black="#000000", lightorange="#E69F00", lightblue="#56B4E9", lightgreen="#6C8645", yellow="#F0E442", blue="#0072B2", orange="#D55E00", pink="#CC79A7",gray="#999999")
```

Once again, this lab is more about statistics than coding *per se*.

We will look at:

- One categorical variable: Chi-square goodness of fit tests
- Two categorical variables: Risk ratio, Odds Ratio,Chi-square independence test; Fisher’s test*

* Lecture slides for these on the course website.



## One categorical variable: $\chi^2 $ goodness of Fit tests

There are many functions associated with the $\chi^2$ distribution in R:



- `dchisq`: returns the value of the Chi-Square probability density function.
- `pchisq`: returns the value of the Chi-Square cumulative density function.
- `qchisq`: returns the value of the Chi-Square quantile function.
- `rchisq`: generates a vector of Chi-Square distributed random variables.



### Proportional Model: NHL example

We looked at this example when discussing the proportional model and its use in a goodness-of-fit test. Our hypotheses were:

$H_0$: Elite hockey players are born randomly across the year.

$H_A$: Elite hockey players are NOT born randomly across the year.

```{r nhl}
# nhl players' birthdays
nhl <- tibble( month = c("Jan", "Feb", "Mar", "Apr", "May", "June", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
        month.num = 1:12,
        births = c(133, 125, 114, 119, 119, 123, 96, 91, 83, 84,73,85)) 

# turn month into ordered factor (ordered by month.num). For plots
nhl<-nhl |> 
  mutate(month = fct_reorder(.f = month, .x = month.num,.desc = TRUE))
nhl
```

Let's start by making a barplot of these NHL births (one categorical variable):

```{r nhl-plot}

ggplot(data = nhl, aes(x = month, y = births)) + 
  geom_col(fill=okabe_ito['orange']) 


```

# now let's make this plot a little prettier

```{r nhl-plot2}
library(ggpubr) #pretty ggplot themes

nhl.plot<-ggplot(data = nhl, aes(x = month, y = births)) + 
  geom_col(fill=okabe_ito['orange']) +
  coord_flip()+ #flip cartesian coords.
  annotate(geom = "text", x = 12:1, y = 5,label = nhl$births, color = "white" ) + #add numbers to bars
  ggtitle("Number of NHL players born by month") + #add title
  theme_minimal() + #pretty theme
  theme(axis.ticks = element_blank(), #remove axis tics
  axis.title.x = element_text(
      size = 14, #larger x title
      hjust = 1 #right align x title
    ),
  axis.title.y = element_text(
    size=14, #larger y title
     hjust = 1 #right align x title
    )
  )
  
nhl.plot  # better!

```

Next thing we need is our expected number of NHL births per each month. This comes from a giant dataset of births by month from the general population.[^1]

```{r births}

# human births per month

births<-tibble( month = c("Jan","Feb","Mar","Apr","May",
                            "June","Jul","Aug","Sep","Oct","Nov","Dec"),
`births (%)`  = 100*c(Jan = 0.0794, Feb = .0763, Mar = .0872,
Apr = .0863, May = .0895, June = .0857, Jul = .0876,
Aug = .085, Sep =  .0854, Oct = .0819, Nov = .0770,
Dec = .0787) )

```

If you don't know this already, here's a neat way to format a table/data.frame/tibble in a pretty way in your knitted document:

[^1]: Ventura et al. (2001). Births: final data for 1999. National Vital
Statistics Reports Vol. 49, No. 1.

```{r births-kable}
library(kableExtra)
births |> kable() # there are many ways to stylize this and customize it
```

Let's add expected births per month to the NHL dataset:


```{r nhl-exp}
# I will do this by joining the NHL and births datasets by a column they have in common: month. I will use dplyr::inner_join()
nhl<-full_join(nhl, births) |> #add a column with expected prop
  mutate(expected.prop=`births (%)`/100) |> #convert % to prop
  mutate(expected.n = expected.prop * sum(births), #expected.n
         `expected %` = 100 * expected.prop) #expected %
nhl
```

Now we are ready to conduct a $\chi^2$ goodness-of-fit test. 

Let's use `mutate` to create a new column with $\frac{(O-E)^2}{E}$:

```{r nhl-chisq}
nhl<-nhl |> #(O-E)^2)/E for each row
  mutate(`sq_dev_over_expect` = (expected.n - births)^2 / expected.n) 
#nhl |> kable() # have a look

# get chi-squared by adding all values in the column
obs.chi2 <-  nhl |> summarise( 
                       chi2 = sum(`sq_dev_over_expect`))      
obs.chi2
```

Now we can use `pchisq` to find the p-value.

```{r nhl-pchisq}
# first define degrees of freedom: 
degF<-nrow(nhl)-1-0
# pull chi2 value from tibble into a vector
obs.chi2<-pull(obs.chi2)

#pchisq: 
pchisq(q = obs.chi2, df=11,lower.tail = F)  # a very low p-value

```

Just for fun, let's plot what the probability density for `df=11` looks like:


```{r dchisq}
#simple base R plot
curve(dchisq(x, df=11), from=0, to=100)


```

Another way of obtaining this p-value is using the `chisq.test` function. It expect as input either a numeric vector or a matrix. Let's do both.


```{r nhl-chisq.test}
# with  numeric vector x with observed births and p=exp.prop for each month
chisq.test(x = nhl$births, p = nhl$expected.prop)
```


### Binomial Model: Monkey Flower example

### Poisson Model: Lousy Fish example


```{r fish1, echo=T, eval=T}


#create tibble with data
lousy.fish <- tibble(num_parasites = 0:9, num_fish = c(103, 72, 44, 14, 3, 1, 1,0,0,0)) 
lousy.fish
# create totals (fish and parasites)
fish.sum   <- lousy.fish |> summarise(total_fish = sum(num_fish), 
                                       total_parasites = sum(num_fish* num_parasites))
fish.sum
```

Make a pretty table for your knitted document:


```{r fish2}
lousy.fish |> kable() |> kable_styling(font_size = 16)

```



## Relative Risk

We will use the titanic dataset for the RR and OR examples.

```{r titanic}
titanic <- read.csv(
  url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter09/chap09f01Titanic.csv"), 
  stringsAsFactors = FALSE)
head(titanic)
titanicTable <- table(titanic$survival, factor(titanic$gender, levels=c("Women", "Men")), useNA = c("ifany"))
titanicTable

# mosaic plot
mosaicplot( t(titanicTable), col = c("firebrick", "goldenrod1"), cex.axis = 1, 
    sub = "Gender", ylab = "Relative frequency", main = "")
```


The relative risk (RR) is the probability of the undesired outcome (death) in the treatment group (women) divided by the probability in the control (men) group.


Our contingency table titanicTable has the undesired outcome (death) in the treatment (women) group in the top left position. The `riskratio()` function from epitools expects the opposite layout, so we need to flip the table and reverse the order of the rows to use it. We can do this all at once with the following arguments to the riskratio() function. The full output of the command is shown below.


```{r}
library(epitools)

# Using the wald method for the CI
RR1 <- riskratio(t(titanicTable), rev = "both", method = "wald")
RR1
#To grab the key output, namely the estimate of the RR and the 95% confidence interval, use the following command:

RR1$measure[2,]

# here is the same thing but using a different method to obtain the CIs:
# using bootstraps for the CI
RR2<-riskratio(t(titanicTable), rev = "both", method = "boot")
RR2
RR2$measure[2,]
```


## Odds Ratio

In R, the simplest way to estimate an odds ratio is to use the command fisher.test(). This function will also perform a Fisher’s exact test (more on that later). The input to this function is a contingency table.


```{r OR}

# OR

OR1<-fisher.test(titanicTable)

# you can use $ to get the values of interest
# OR estimate
OR1$estimate
# CI for the OR
OR1$conf.int

```

Alternatively, you can use the package `epitools`.

```{r epiOR}
library(epitools)

# using fisher method for the OR and CI. this gives the same result as fisher.test()
OR3<-oddsratio(titanicTable, method = "fisher")
OR3
OR3$measure[2,]

```

## Toxoplasma example

```{r toxop}
toxoplasma <- read.csv(
  url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter09/chap09e3ToxoplasmaAndAccidents.csv"), stringsAsFactors = FALSE)
head(toxoplasma)
```

First, make the 2 x 2 contingency table. The table that results has the undesired outcome (accidents) for the treatment group (infected) in the top left.

```{r toxop1}
toxTable <- table(toxoplasma$accident, toxoplasma$condition)
toxTable
# let's also make a mosaic plot
mosaicplot(toxTable, col=c("firebrick", "goldenrod1"), cex.axis = 1.2, 
sub = "Condition", dir = c("h","v"), ylab = "Relative frequency", main = "")
```


OR using epitools:


```{r ortoxop2}
OR <- oddsratio(toxTable, method = "fisher")
OR$measure[2,]
```

