---
title: "4.2.Estimation with Uncertainty"
author: "Bárbara D. Bitarello"
date: "10-27-2025"
keywords: "Estimation"
format:
  revealjs:
    embed-resources: false
    smaller: true
    scrollable: true
    incremental: true
    slide-number: true
    transition: fade
    background-transition: fade
    highlight-style: ayu-dark
    footer: "B215: Biostatistics with R"
    center-title-slide: false
    theme: [custom.scss]
    callout-appearance: simple
    code-fold: true
    auto-slide: 133000
code-link: true
execute:
  echo: false
  freeze: false
knitr: 
  opts_chunk:
    tidy: true
    tidy-opts: { "value": "list(width.cutoff = I(60))", "tag": "!expr" }
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=T)
library(tidyverse)
library(forcats)
library(dplyr)
library(ggplot2)
library(ggforce)
library(tidyr)
library(stringr)
library(knitr)
library(ggthemes)
library(wesanderson)
library(gt)
library(RColorBrewer)
library(plotly)
library(readr)
library(thematic)
library(infer)
library(kableExtra)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
source("bb_theme.R")

# set the theme
theme_set(bb_theme())
#theme_set(theme_bw())
#thematic_rmd(font = "Roboto Condensed", qualitative = asteroidcity1())
set.seed(12345)

tigerData<-read.csv("~/Documents/GitHub/DataSetsb215/data-raw/chap02e2aDeathsFromTigers.csv")

DieRoll<-function(times = 1){
  sample(x = 1:6,size = times, replace = T)
}

## to fix annoying ggplot2 issue ----
##https://rstudio-pubs-static.s3.amazonaws.com/991178_746a3f129f4d46829bda51127a8cb8f7.html
StatCount2 <- ggproto(
  "StatCount2", StatCount,
  compute_panel = function (self, data, scales, ...) {
    if (ggplot2:::empty(data)) 
      return(ggplot2:::data_frame0())
    groups <- split(data, data$group)
    stats <- lapply(groups, function(group) {
      self$compute_group(data = group, scales = scales, ...)
    })
    non_constant_columns <- character(0)
    
    stats <- mapply(function(new, old) {
      if (ggplot2:::empty(new)) return(ggplot2:::data_frame0())
      # Edit 1: Identify exceptions where a variable is constant *within values of x*
      within_x_constant <- vapply(old, function(x) nlevels(interaction(x, old$x, drop = TRUE)) == max(old$x), logical(1L))
      # Edit 2: Columns to keep
      kept <- unique(old[, within_x_constant])
      # Edit 3: Don't test non-constant-ness of `kept` columns
      old <- old[, !(names(old) %in% c(names(new), names(kept))), drop = FALSE]
      non_constant <- vapply(old, function(x) length(unique0(x)) > 1, logical(1L))
      non_constant_columns <<- c(non_constant_columns, names(old)[non_constant])
      # Edit 4: Add `kept` columns
      vctrs::vec_cbind(
        new,
        kept[, setdiff(names(kept), "x"), drop = FALSE],
        old[rep(1, nrow(new)), setdiff(names(old), names(kept)), drop = FALSE],
      )
    }, stats, groups, SIMPLIFY = FALSE)
    
    non_constant_columns <- ggplot2:::unique0(non_constant_columns)
    dropped <- non_constant_columns[!non_constant_columns %in% self$dropped_aes]
    if (length(dropped) > 0) {
      cli::cli_warn(c(
        "The following aesthetics were dropped during statistical transformation: {.field {glue_collapse(dropped, sep = ', ')}}",
        "i" = "This can happen when ggplot fails to infer the correct grouping structure in the data.",
        "i" = "Did you forget to specify a {.code group} aesthetic or to convert a numerical variable into a factor?"
      ))
    }

    # Finally, combine the results and drop columns that are not constant.
    data_new <- ggplot2:::vec_rbind0(!!!stats)
    data_new[, !names(data_new) %in% non_constant_columns, drop = FALSE]
  }
)

n.stars <- c(n_gold_small  =14*3,
             n_gold_medium = 8*3,
             n_gold_large  = 6*3,
             n_blue_small  = 6*3,
             n_blue_medium = 8*3,
             n_blue_large  = 3*3,
             n_red_small   = 7*3,
             n_red_medium  = 8*3,
             n_red_large   = 0*3)
stars <- tibble(color = rep( (str_extract(names(n.stars), "n_..*_") |> 
                    str_replace_all(pattern = "n|_","")),  times= n.stars), 
                size = rep((str_replace(names(n.stars), "n_","") )  |> 
                             str_extract( "_..*") |> 
                             str_replace_all(pattern = "_",""),  times = n.stars)) |>
  sample_n(size = sum(n.stars), replace = FALSE)  |>
  mutate(x = rep(1:14, times = 14)[1:sum(n.stars)],
         y = rep(1:14, each = 14)[1:sum(n.stars)],
         size = fct_relevel(size, "small","medium","large"))

human_genes<-read_csv("~/Documents/GitHub/Matb215f2025/Rlabs/lab6/human_genes.csv") |> filter(size <=15000)
```

## Outline

-   ~~Sampling - what’s the point?~~

-   ~~Getting a feel for sampling~~

-   The sampling distribution - there's MORE!

-   Standard error

-   Standard error of the mean

To be continued:

-   Confidence intervals

-   CI of the mean

-   Error bars - common mistakes

-   Pseudo-replication

## Recap question

Which of the following describes the sampling distribution of an
estimate?

A. The set of all values you get in a given sample from a population.

B. The set of all values you might get when you sample a population.

C. The set of all values in a given population.

D. The set of all values in a given sample.

## Recap Question

Which of the following describes the sampling distribution of an
estimate?

A. The set of all values you get in a given sample from a population.

**B. The set of all values you might get when you sample a population.**

C. The set of all values in a given population. 

D. The set of all values in a given sample.

## Recap Question 2

The standard deviation of an estimate's sampling distribution is called which of the following?

A. Standard distribution.

B. Standard error.

C. Standard probability distribution.

D. Standard variation.

## Recap Question 2

The standard deviation of an estimate's sampling distribution is called which of the following?

A. Standard distribution.

**B. Standard error.**

C. Standard probability distribution.

D. Standard variation.

## In the face of chance (randomness of my sample), can I trust my estimate? And how much can I trust it? {.dark-theme .center}

## Two key measures of precision of an estimate

-   Standard Error

-   Confidence Intervals

## The Standard Error: Definition

-   The standard error reflects the difference between an estimate and
    the target parameter value.

-   The standard error predicts the sampling error of the estimate.

-   The standard error of an estimate is the standard deviation of its
    sampling distribution.
    
- We could in principle calculate SE for any summary statistic but we will focus on the most common -- the standard error of the mean.

## The Standard Error of the Mean (SEM)

The SEM quantifies how precisely you know the population mean.

-   Because we rarely know the population standard
    deviation,<font color = #006475>$\sigma$</font>, we cannot find the
    parameter <font color = #006475>
    $\sigma_{\bar{Y}}=\frac{\sigma}{\sqrt{n}}$</font> , the standard
    error of the <font color = #006475>population mean</font>.

-   But, we can use the sample standard deviation,
    <font color = #E69F00>$s$</font>, to estimate
    <font color = #E69F00>$SEM_{\bar{Y}}=\frac{s}{\sqrt{n}}$</font>, the
    standard error of the <font color = #E69F00>sample mean</font>.

- We can also estimate $SEM_{\bar{Y}}$ by taking the standard deviaton of the sampling distribution of sample means of a given size.

::: aside
Note we use SEM instead of SE to avoid ambiguities. Standard errors can
be calculated for other statistics besides the mean.
:::

## Example: human gene lengths {.dark-theme .center}

## Human gene lengths

```{r geneleng, fig.cap="Note: parameters shown are for full dataset (n=20,290) but 26 genes with length > 15,000 \nwere omitted from this plot.", tidy="styler"}
human.hist<-ggplot(human_genes , aes(x=size)) + geom_histogram(aes(y=after_stat(count / sum(count))),bins=60, color="lightgray", fill="#006475") + 
  xlab("Gene length (basepairs)") + 
  ylab("Relative Frequency")+
  ggtitle(stringr::str_wrap("Distribution of human gene lengths (aka, the population of gene lengths)", width = 40))+
  annotate('text', x = 10000, y = 0.04, 
        label = "mu==3511.5",parse = TRUE,size = 8) +
  annotate('text', x = 10000, y = 0.03, 
        label = "sigma==2833.3",parse = TRUE,size = 8)+
  bb_theme()
human.hist
```

## A random sample of five genes

```{r geneleng2, echo = F, eval=T}
set.seed(1234)
library(gt)

samp5<-human_genes |> 
  slice_sample(n = 5, replace = F) |>
  select(size)
samp5
```
## A random sample of five genes

Summarising my sample of $n=5$:

```{r geneleng3, echo = F, eval=T}

samp5 |> 
  summarise(Mean=mean(size), SD=sd(size)) |> 
  gt()
```


## Another random sample of five genes

Summarising my other sample of $n=5$:

```{r geneleng4, echo = F, eval = F}
set.seed(1235)

samp5_2<-human_genes |> 
  slice_sample(n = 5, replace = F) |>
  select(size)
samp5_2 |> 
  summarise(Mean_length=mean(size), SD_length=sd(size))
```

```{r geneleng5, echo = F}
set.seed(1235)

samp5_2<-human_genes |> 
  slice_sample(n = 5, replace = F) |>
  select(size)
samp5_2 |> gt()

samp5_2 |> 
  summarise(Mean_length=mean(size), SD_length=sd(size)) |>
  gt()
```

## I can start combining those different estimates

```{r geneleng6, reps, echo = F}
#bind rows using bind_rows

samp.dist.partial<-bind_rows(samp5 |> 
    summarise(Mean_length=mean(size), 
              SD_length=sd(size)),
    samp5_2 |> 
    summarise(Mean_length=mean(size), 
              SD_length=sd(size))
)
    
   
```

```{r geneleng7, echo = F}
samp.dist.partial |> gt()
```

## Ten samples of five 5 genes


```{r sampdist5, tidy="styler"}
library(infer)
set.seed(1985)
samp.dist5<-human_genes |>
  rep_sample_n(size = 5, replace = FALSE,reps = 10) |>
  summarise(mean_length= mean(size), 
            sd_length = sd(size))
```

```{r sampdist6, echo = F}
samp.dist5 |> gt()
```

```{r sampdistsumm, tidy="styler"}
genes.sum <- samp.dist5|> 
  ungroup() |> 
  summarise(mean_of_means = mean(mean_length), 
            sd_of_means = mean(sd_length), 
            SEM = sd(mean_length)) |> 
  unlist() |> 
  round(digits = 2)
```

```{r sampdist7, echo = F}
genes.sum<-as_tibble(t(c(genes.sum)))
genes.sum |> mutate(Real_SEM=sd(human_genes$size)) |> gt()
```

::: aside
I will make the code for these simulations available
:::

## Many

```{r sampdist5_2, tidy="styler"}
library(infer)
set.seed(1985)
samp.dist5<-human_genes |>
  rep_sample_n(size = 5, replace = FALSE,reps = 10000) |>
  summarise(mean_length= mean(size), 
            sd_length = sd(size))
genes.sum <- samp.dist5|> 
  ungroup() |> 
  summarise(mean_of_means = mean(mean_length), 
            sd_of_means = mean(sd_length), 
            SEM = sd(mean_length)) |> 
  unlist() |> 
  round(digits = 2)
```
### Sampling Distribution

```{r sampdist8, tidy="styler", fig.height=4}
genes.sum.plot <- ggplot(samp.dist5, aes(x = mean_length)) +
  geom_histogram(aes(y=after_stat(count)/sum(count)), bins = 100, color = "lightgray", fill = "#E69F00") +
  ggtitle(stringr::str_wrap("Sampling distribution of mean gene lengths", width = 40), subtitle= "(n = 5; 10,000 replicates)")+
  xlab(label = "Mean length (basepairs)")+
  ylab("Proportion of samples")+
   annotate(geom = "text", x = 4500, y =.02, label =sprintf( "mean of means = %s" ,genes.sum[1]), size = 5, hjust = 0) +
  geom_vline(xintercept = mean(human_genes$size), color = "white", lty = 2) +
  bb_theme() 
genes.sum.plot 

#annotate(geom = "text", x = 4500, y =.025, label =sprintf( "mean mean = %s\n mean sd = %s",genes.sum[1], genes.sum[2]), size = 5, hjust = 0)+
```


::: aside
More formally, the "mean of the means" is the "grand mean", not the "mean of means" (my favorite). (-:
:::

## Compare the pop & the sample dist

```{r sampdist9,echo=FALSE,warning=FALSE, message=FALSE, fig.align='center'}
library(gridExtra)
grid.arrange(human.hist, genes.sum.plot,nrow = 1)
```

-   **Mean** parameter estimates of a sampling distribution should equal
    population parameters.


## Compare the pop & the sample dist

```{r sampdist10,echo=FALSE,warning=FALSE, message=FALSE, fig.align='center'}
library(gridExtra)
genes.sum.plot2<-genes.sum.plot+
  annotate(geom = "text", x = 4500, y =.018, label =sprintf( "sd of means = %s", genes.sum[3]), size = 5, hjust = 0)
grid.arrange(human.hist, genes.sum.plot2,nrow = 1)
```

-   **The width** (standard deviation) of a population ($\sigma=2833.3) exceeds that of its sampling distribution ($s_{\text{all means}}=1277.22)


## The Standard Error & The Sampling Distribution

-   The standard error is the standard deviation of the sampling
    distribution 

-   The standard deviation of this distribution is $1277.22$ basepairs

- So $1277.22$ is our **estimate** for the standard error of the mean based on our sampling distribution

```{r sem, echo=F,fig.align='center'}
genes.sum.plot2 <- ggplot(samp.dist5, aes(x = mean_length)) +
  geom_histogram(aes(y=after_stat(count)/sum(count)), bins = 100, color = "lightgray", fill = "#E69F00") +
  ggtitle(stringr::str_wrap("Sampling distribution of human gene lengths", width = 40), subtitle= "(n = 5, 10,000 replicates)")+
  xlab(label = "Mean length (basepairs)")+ylab("Proportion of samples")+
  annotate(geom = "text", x = 4500, y =.025, label =sprintf( "mean mean = %s\n mean sd = %s\n sd mean",genes.sum[1], genes.sum[2], genes.sum[3]), size = 5, hjust = 0)+
  geom_vline(xintercept = mean(human_genes$size), color = "white", lty = 2) +
  bb_theme() 
genes.sum.plot2 
```

## The Standard Error & The Sampling Distribution

```{r sampdist11, fig.align='center'}
genes.sum.plot2
```

-   You might be thinking, but how would I get such a sampling
    distribution?

-   It turns out there is a clever way of resampling from your own
    sample to estimate an (unknown) sampling distribution!

## Bootstrapping! 

-   use the observed sample to estimate the population distribution.

-   then samples can be drawn from the estimated population and the
    sampling distribution of any type of estimator can itself be
    estimated.

[![From:
https://online.stat.psu.edu/stat555/node/119/](images/clipboard-757051427.png){fig-alt="Image explaining bootstrap algorithm"
width="414"}](https://online.stat.psu.edu/stat555/node/119/)

::: aside
More on this later
:::

## Uncertainty

Because we rarely know the <font color = #006475 >
population</font> sd $(\sigma)$, we cannot find the <font color = #006475 >parameter
$\sigma_{\bar{Y}}= \sigma_Y / \sqrt{n}$ </font>.

But, we can use the <font color = #E69F00>sample</font>
standard deviation ($s$) to estimate the standard error
<font color = #E69F00>$SEM_{\bar{Y}}= s / \sqrt{n}$</font>

## Compare

The sampling distribution estimate of $SEM_{\bar{Y}}=1277.22$

and

the parameter Standard Error of the population mean

$\sigma_{\bar{Y}}=\frac{\sigma_{Y}}{\sqrt{n}}$

$=\frac{2833.2}{\sqrt{5}}=1267.046$

::: goal
These are indeed very similar! $SEM_{\bar{Y}}$ is an **estimate** of
$\sigma_{\bar{Y}}$ and is itself estimated with error.
:::

## Sample size, uncertainty, and the standard error {.dark-theme .center}

## SE Decreases with Sample Size

The standard error goes down as a sample size goes up.

Using the example from before:

$\sigma_{\bar{Y}}=\frac{\sigma_{Y}}{\sqrt{n}}$

if $n=5$

$\sigma_{\bar{Y}}=\frac{2833.3}{\sqrt{5}}=1267.09$

if $n=100$

$\sigma_{\bar{Y}}=\frac{2833.3}{\sqrt{100}}=283.33$

## Small Samples, Large Uncertainty

-   Uncertainty increases as sample size decreases.

-   Therefore, extreme values are often associated with small sample
    size.

::: goal
Treat simple summary statistics with skepticism.

Be sure to consider uncertainty before being misled by exceptional
values.
:::

## Standard error - Summary {visibility="hidden"}

::: goal
-   Estimates with smaller standard errors are more precise
-   The smaller the sampling error, the less uncertainty there is about
    the target parameter in the population (“out there”).
-   You can calculate the standard error for any estimate, not just the
    mean.
-   The SE of the sample mean if sometimes referred to as SE, but this
    is ambiguous - use SEM instead of SE
-   For huge sample sizes, the SEM is always tiny, even if the data have
    considerable spread (SD).
-   Small samples underestimate the SD of the population
-   Both SD and SEM are expressed in the same unit as the data
:::

## Simulators to help you grasp this:

- https://shiny.abdn.ac.uk/Stats/apps/app_sampling/
- https://www.zoology.ubc.ca/~whitlock/Kingfisher/SamplingNormal.htm
- http://shiny.calpoly.sh/Sampling_Distribution/

## That's all for today

![From:
makeameme.org](images/forrestgump.png){fig-alt="\"Forrest says \"And that's all I wanted to say about that\""
width="347"}
