---
title: "4.5.Estimation with Uncertainty"
subtitle: "Confidence Intervals"
author: "Bárbara D. Bitarello"
date: "11-5-2025"
keywords: "Confidence Intervals & Pseudoreplication"
format:
  revealjs:
    embed-resources: false
    smaller: true
    scrollable: false
    incremental: false
    slide-number: true
    transition: slide
    background-transition: fade
    highlight-style: ayu-dark
    footer: "B215: Biostatistics with R"
    center-title-slide: false
    theme: [custom.scss]
    callout-appearance: simple
    code-fold: true
    chalkboard: true
    auto-slide: 133000
code-link: true
execute:
  echo: false
  freeze: false
knitr: 
  opts_chunk:
    tidy: true
    tidy-opts: { "value": "list(width.cutoff = I(60))", "tag": "!expr" }
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=T)
library(tidyverse)
library(forcats)
library(dplyr)
library(ggplot2)
library(ggforce)
library(tidyr)
library(stringr)
library(knitr)
library(ggthemes)
library(wesanderson)
library(gt)
library(RColorBrewer)
library(plotly)
library(readr)
library(thematic)
library(infer)
library(kableExtra)
library(gridExtra)


gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
source("bb_theme.R")

# set the theme
theme_set(bb_theme())
#theme_set(theme_bw())
#thematic_rmd(font = "Roboto Condensed", qualitative = asteroidcity1())

## to fix annoying ggplot2 issue ----
##https://rstudio-pubs-static.s3.amazonaws.com/991178_746a3f129f4d46829bda51127a8cb8f7.html
StatCount2 <- ggproto(
  "StatCount2", StatCount,
  compute_panel = function (self, data, scales, ...) {
    if (ggplot2:::empty(data)) 
      return(ggplot2:::data_frame0())
    groups <- split(data, data$group)
    stats <- lapply(groups, function(group) {
      self$compute_group(data = group, scales = scales, ...)
    })
    non_constant_columns <- character(0)
    
    stats <- mapply(function(new, old) {
      if (ggplot2:::empty(new)) return(ggplot2:::data_frame0())
      # Edit 1: Identify exceptions where a variable is constant *within values of x*
      within_x_constant <- vapply(old, function(x) nlevels(interaction(x, old$x, drop = TRUE)) == max(old$x), logical(1L))
      # Edit 2: Columns to keep
      kept <- unique(old[, within_x_constant])
      # Edit 3: Don't test non-constant-ness of `kept` columns
      old <- old[, !(names(old) %in% c(names(new), names(kept))), drop = FALSE]
      non_constant <- vapply(old, function(x) length(unique0(x)) > 1, logical(1L))
      non_constant_columns <<- c(non_constant_columns, names(old)[non_constant])
      # Edit 4: Add `kept` columns
      vctrs::vec_cbind(
        new,
        kept[, setdiff(names(kept), "x"), drop = FALSE],
        old[rep(1, nrow(new)), setdiff(names(old), names(kept)), drop = FALSE],
      )
    }, stats, groups, SIMPLIFY = FALSE)
    
    non_constant_columns <- ggplot2:::unique0(non_constant_columns)
    dropped <- non_constant_columns[!non_constant_columns %in% self$dropped_aes]
    if (length(dropped) > 0) {
      cli::cli_warn(c(
        "The following aesthetics were dropped during statistical transformation: {.field {glue_collapse(dropped, sep = ', ')}}",
        "i" = "This can happen when ggplot fails to infer the correct grouping structure in the data.",
        "i" = "Did you forget to specify a {.code group} aesthetic or to convert a numerical variable into a factor?"
      ))
    }

    # Finally, combine the results and drop columns that are not constant.
    data_new <- ggplot2:::vec_rbind0(!!!stats)
    data_new[, !names(data_new) %in% non_constant_columns, drop = FALSE]
  }
)

human_genes<-read_csv("~/Documents/GitHub/Matb215f2025/Rlabs/lab6/human_genes.csv")

DieRoll<-function(times = 1){
  sample(x = 1:6,size = times, replace = T)
}
CoinToss<-function(times=1){
  sample(x=c("heads", "tails"), size=times, replace=T)
}
MakeSampleMeansGraph <- function(samples, actual.mean,x_lim = NA,...){
  if(is.na(x_lim)){x_lim <-range(c(samples))}
  sample.means <- colMeans(samples)
  sample.SEs   <- apply(samples,2,sd)/sqrt(nrow(samples))
	plot(1,xlim = x_lim , ylim = c(0,(1+ncol(samples))),
		type="n",axes = FALSE, ylab = "",...)
	segments(x0 = (sample.means - 2 * sample.SEs) , 
           y0 = seq_along(samples[1,]), 
           x1 = (sample.means + 2 * sample.SEs),  
           y1 =  seq_along(samples[1,]), 
           col="blue" )
	abline(v=actual.mean, col = "red")
	box()
	axis(1)
	mtext("sample", side =2 ,line = .5)
}
MakeSampleMeansGraph2 <- function(samples, actual.mean,title="title", x_lim = NA,...){
  if(is.na(x_lim)){x_lim <-range(c(samples$size))}
  sample_means<-samples |> summarise(sample_mean=mean(size), sample_sd=sd(size), n=n(),sem=sample_sd/sqrt(n), CI_U=sample_mean+2*sem, CI_L=sample_mean-2*sem)
qtext<-stringr::str_wrap(paste0(sample_means |> group_by(Logic=actual.mean >= CI_L & actual.mean <=CI_U) |> tally() |> filter(Logic==TRUE) |> pull(n), "/", max(samples$replicate),  " CIs include population mean"),16)
 
p1<-ggplot(sample_means, aes(x = sample_mean, y=replicate))+
   geom_point(color="#D55E00")+
   geom_errorbar(aes(xmin=CI_L, xmax=CI_U)) + 
   ylab("sample") +
   xlab(stringr::str_wrap("Mean gene length (basepairs)",20)) + 
   geom_vline(xintercept=actual.mean,colour = "#0072B2", lty=2,linewidth = 1)+
  annotate(geom = "text",x=actual.mean+2000,y = 103, label="Population mean",colour = "#0072B2")+
  annotate(geom = "text", x = 12000, y = 50, label=qtext)+
  #xlim(0, 15000)+
  ggtitle(title)+
   bb_theme()+
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
}


MakeSampleMeansGraph3 <- function(samples, actual.mean,title="title", x_lim = NA,...){
  if(is.na(x_lim)){x_lim <-range(c(samples$size))}
  sample_means<-samples |> group_by(replicate) |> summarise(sample_mean=mean(size), sample_sd=sd(size), n=n(),sem=sample_sd/sqrt(n), CI_U=sample_mean+2*sem, CI_L=sample_mean-2*sem)
qtext<-stringr::str_wrap(paste0(sample_means |> group_by(Logic=actual.mean >= CI_L & actual.mean <=CI_U) |> tally() |> filter(Logic==TRUE) |> pull(n), "/", max(samples$replicate),  " CIs include population mean"),16)
 
p1<-ggplot(sample_means, aes(x = sample_mean, y=replicate))+
   geom_point(colour = "#D55E00")+
   geom_errorbar(aes(xmin=CI_L, xmax=CI_U), color="#616571") + 
   ylab("sample") +
   xlab(stringr::str_wrap("Mean gene length (basepairs)",20)) + 
   geom_vline(xintercept=actual.mean,colour = "#0072B2", lty=2,linewidth = 1)+
  annotate(geom = "text",x=actual.mean+2000,y = 103, label="Population mean",colour = "#0072B2")+
  annotate(geom = "text", x = 12000, y = 50, label=qtext)+
  xlim(0, 15000)+
  ggtitle(title)+
   bb_theme()+
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
}
```

## Pratice question 1

If the 95% confidence interval for a data set is $15 < \mu < 23$, which
of the following are closest to the standard deviation and sample size
for the data set? Hint: use the 2SEM rule of thumb.

a.  [ ] Standard deviation = 8, sample size = 4
b.  [ ] Standard deviation = 8, sample size = 16
c.  [ ] Standard deviation = 16, sample size = 4
d.  [ ] Standard deviation = 16, sample size = 16

## Practice question 1

If the 95% confidence interval for a data set is $15 < \mu < 23$, which
of the following are closest to the standard deviation and sample size
for the data set? Hint: use the 2SEM rule of thumb.

a.  [ ] Standard deviation = 8, sample size = 4 SEM=4
b.  [x] Standard deviation = 8, sample size = 16 *SEM = 2*
c.  [ ] Standard deviation = 16, sample size = 4 SEM=8
d.  [ ] Standard deviation = 16, sample size = 16 -\> SEM=4

Sample mean is in the middle of the interval, so 23-15=8; 23-4=19;
\$\bar x = 19 2SEM rule: Lower: $19 - 2\times SEM_{\bar x}=15$;
$4=2\times SEM_{\bar x}$; $\frac{s}{\sqrt(n)}=2$ Can't be

## Practice Question 2

What is wrong with this statement?

> Measurements of the distance between the canine tooth and last molar
> for 35 wolf upper jaws were made by a researcher. They found the 95%
> confidence interval for the mean to be $10.16 < \mu < 10.47$ and the
> 99% confidence interval to be $10.21<\mu<10.44$.

## Practice Question 2

What is wrong with this statement?

> Measurements of the distance between the canine tooth and last molar
> for 35 wolf upper jaws were made by a researcher. They found the 95%
> confidence interval for the mean to be $10.16 < \mu < 10.47$ and the
> 99% confidence interval to be $10.21<\mu<10.44$.

These cannot both be true because for a given sample the 99% CI MUST BE
wider than the 95% CI.

::: aside
This was taken from a previous exam in Biol 215...
:::

## CI Web Apps:

I highly recommend using these:

-   Web app from the book:
    <http://www.zoology.ubc.ca/~whitlock/Kingfisher/CIMean.htm> Note:
    this uses the method involvign the Normal distribution. If you click
    on calculations you will see it. We will learn that later.

-   And this one: <https://shiny.abdn.ac.uk/Stats/apps/app_sampling/>
    This allows you to do something very similar as to what I did in the
    slides, step by step.

## Let's pratice 3 methods of obtaining CI {visibility="hidden"}

::: goal
For a random sample of $n=100$ genes from the human_genes dataset,
$\bar Y=3489$, $s=2238$.
:::

**1)** Based on the normal distribution (we won't learn to calculate
this yet, but just for comparison):

$\bar Y \pm Z \times \frac{\sigma}{\sqrt(100)}$

$\left[2933.7, 4044.3\right]$

**2a) Rule of thumb:** ***if*** we knew the true standard deviation
($\sigma=2833.293$)

**2b) Rule of thumb:** if we did *not* know the true standard deviation
(more likely)

## Let's pratice 3 methods of obtaining CI {visibility="hidden"}

::: goal
For a random sample of $n=100$ genes from the human_genes dataset,
$\bar Y=3489$, $s=2238$.
:::

::::::: columns
:::: {.column width="50%"}
**2a) Rule of thumb:** ***if*** we knew the true standard deviation
($\sigma=2833.293$)

::: incremental
$\bar x \pm (2\times \frac{2833.293}{\sqrt(100)}$

$3489 \pm 2\times 283.3293 = 3489 \pm 566.6586$

Upper 95% CI: $3489+566.6586=4055.6$

Lower 95% CI: $3489-566.6586=2922.3$

$\left[2922.3,  4055.6\right]$
:::
::::

:::: {.column width="50%"}
**2b) Rule of thumb:** if we did *not* know the true standard deviation
(more likely), but other assumptions were met (large enough sample size,
pop variable of interest normally distributed).

::: incremental
$\bar x \pm (2\times \frac{2238}{\sqrt(100)}$

$3489 \pm 2\times 223.8=3489 \pm 447.7$

Upper 95% CI: $3489+447.7=3936.7$

Lower 95% CI: $3489-447.7=3041.3$

$\left[3041.3,3936.7\right]$
:::
::::
:::::::

## CIs from bootstrapping

This is a great choice because it does not make any assumptions about
underlying distributions.

::: goal
Recipe for non-parametric bootstrapping:

-   Take a sample of given size (in our example, $n=100$ genes)
-   From this sample, take $R$ equally-sized ($n=100$) samples (with
    replacement) and calculate the statistic of interest (mean gene
    length)
-   Repeat many times (e.g. 10,000)
-   The distribution of mean gene lengths approximates the sampling
    distribution nicely.
:::

We will do this in lab.

## Sampling Distribution vs. Bootstrap Distribution {.dark-theme .center}

## Sampling Distribution

::::: columns
::: {.column width="80%"}
[![From: Data Science in R, a gentle introduction, Fig.
9.1](images/clipboard-564726195.png){fig-alt="Figure 9.1: A review of the thought experiment behind a sampling distribution."
width="496"}](https://bookdown.org/jgscott/DSGI/images/samplingdistribution_schematic.png)
:::

::: {.column width="20%"}
> *Of course, if you really could peer into all those parallel
> universes, each with its own sample from the same data-generating
> process, life would be easy.* [^1]
:::
:::::

[^1]: James Scott, *Data Science in R: a gentle introduction*

## But not all is lost!

Even if you can't repeatedly go to the population and sample over and
over again

[![From: Data Science in R, a gentle introduction, Fig.
9.2](images/clipboard-538087364.png){fig-alt="Figure 9.2: A stylized depiction of a bootstrap sampling distribution of an estimator."
width="550"}](https://bookdown.org/jgscott/DSGI/the-bootstrap.html)

## But not all is lost!

**How?** By pretending that the sample itself represents the population,
which allows one to approximate the effect of sampling variability by
resampling (with replacementt) from the original sample. This is
"bootstrap resampling" or, simply, *"bootstrapping"*.

[![From: Data Science in R, a gentle introduction, Fig.
9.2](images/clipboard-538087364.png){fig-alt="Figure 9.2: A stylized depiction of a bootstrap sampling distribution of an estimator."
width="304"}](https://bookdown.org/jgscott/DSGI/the-bootstrap.html)

## Bootstrapping {visibility="hidden"}

```{r, echo = T}
set.seed(1)
sample.100 <- rep_sample_n(human_genes,size = 100,replace=F, reps=10000) |> summarise(meanL=mean(size), ss = 100, sdL=sd(size), semL=sdL/sqrt(ss))
#SEM:sd(sample.100$meanL) 283.7894
sample.100.boots<-rep_sample_n(
  rep_sample_n(human_genes,size = 100,replace=F, reps=1)|> 
    ungroup() |> 
    select(-replicate), 
  size=100, replace=T, reps=10000) |> 
  summarise(meanL=mean(size), ss = 100, sdL=sd(size), semL=sdL/sqrt(ss))
print(paste0("Bootstrap estimate of SEM (as the sd of sample means): ", round(sd(sample.100.boots$meanL),1)))

print(paste0("Lower 95% CI: 3489 - 2 x ", round(sd(sample.100.boots$meanL),1)))

print(c(3489 - (2 * 261.8), 3489 + (2 * 261.8)))


#fake bootstrapping

fake.boots.100<-rep_sample_n(human_genes,size = 100,replace=F, reps=10000) |> summarise(meanL=mean(size), ss = 100, sdL=sd(size), semL=sdL/sqrt(ss))
#SEM:sd(sample.100$meanL) 283.7894
```

In the long run, \~95% of 95% CIs estimated from random samples will
include the population parameter of interest!

## Comparison table {visibility="hidden"}

::: goal
For a random sample of $n=100$ genes from the human_genes dataset,
$\bar Y=3489$, $s=2238$.
:::

| Normal distribution | Rule of thumb (known pop sd) | Rule of thumb (estimated pop sd) |
|----------------------|-------------------------|-------------------------|
| $\left[2933.7, 4044.3\right]$ | $\left[2922.3,  4055.6\right]$ | $\left[3041.3,3936.7\right]$ |

. . .

| Bootstrapping from one sample of $n=100$ | "Bootstrapping": multiple samples of $n=100$ from pop |
|--------------------------------------|----------------------------------|
| $\left[2965.4, 4012.6\right]$ | $\left[2926.5, 4051.5\right]$ |

## Tradeoffs {visibility="hidden"}

-   We want confidence coefficient to be closer to 1.

-   We want the sample size to be as small as possible (but not too
    small). This is a practical issue.

-   We want the CI to be as narrow as possible

-   Think about the following:

-   As we increase the sample estimate, the CI …?

-   As we decrease st. dev, the CI …?

-   As we decrease the confidence level, ($1-\alpha$), the CI …?

-   As we increase sample size …?

## CI: Q&A to answer your burning questions {.dark-theme .center}

## CI: Q&A

**Q: When can I use the rule of thumb that the CI equals the sample mean
plus or minus 2 SDs?**

*Never!* In a Gaussian distribution you expect about 95% of the
individual values to lie within 2 SD of the mean. But the idea of a CI
is to quantify how precisely you know the population mean. For that, you
need the sample size to be taken into account.

**Q:When can I use the rule of thumb that the CI equals the sample mean
plus or minus 2 SEM?**

When $n$ is sufficiently large and the population variable of interest
is normally distributed (approximately). We will come back to this. If
in doubt, don’t use it!!! \[if you do, acknowledge this shortcoming\]

## CI: Q&A {visibility="hidden"}

**Q: Why do I see 95% CIs everywhere?**

Tradition! CIs of 90% and 99% are less common but also often seen in
biology.

A CI of, say, 50% would give a very narrow interval

A CI of 100% would require the entire population and would be rather
pointless

## PSEUDOREPLICATION: WHEN YOUR CONFIDENCE IS UNDESERVED {.dark-theme .center}

## What is pseudoreplication?

Pseudoreplication is the error that occurs when samples are not
independent, but they are treated as though they are.

-   Thus the sample size is smaller than claimed.

-   Pseudoreplication provides a false sense of high certainty.

-   This is not an uncommon issue in published papers.

## Example

::: goal
You are interested in estimating the average pulse rate of mountain
climbers. Since they are hard to find, you decide to take 10
measurements from each climber. You study 6 climbers, so you have 60
measures. What is your sample size (n)?
:::

It's 6. But you can and should use the extra data. How? Take the mean
for each climber, so that you have 6 pulse rates, one for each climber
($n=6$).

::: goal
This is a simple example but, in practice, **pseudoreplication is hard
to identify** and things differ by field
:::

> Pseudoreplication problems arise from the organisational complexity of
> life itself. (Lazic et al. (2018)

## Requirements for genuine replication

The best place to start: think about what are the biological and
experimental units of your and look for field specific discussion on the
topic. E.g., this publication talks about this pseudoreplication in cell
culture and animal experiments:

> *To ensure (...) results can be extrapolated to the population of
> interest, there must also be replication relevant to the hypothesis
> being tested. (...) define the scientific or biological unit of
> interest, which is the entity (i.e., people, animals, and cells) that
> we would like to test a hypothesis or draw a conclusion about. (...)
> to conclude that a drug is better than a placebo, a large number of
> patients are required because the hypothesis is about patients. We
> cannot give Jim the drug and Bob the placebo, take a daily measurement
> for several weeks, and then make general conclusions about the drug's
> efficacy.* [^2]

[^2]: Lazic et al. (2018) *What exactly is ‘N’in cell culture and animal
    experiments?.* PLoS Biology.
    https://doi.org/10.1371/journal.pbio.2005282

::: aside
:::

## A meta moment {visibility="hidden"}

[![From: Nearly half of the studies (46%, 95% CI = 38%–53%) confused
pseudoreplication for genuine replication while 32% (95% CI = 26% to
39%) did not provide enough information to determine if the sample size
was correct (A). Consistent with previous research, randomisation,
blinding, and the sample size was not always reported (B). Error bars
are 95%
CI.](images/pbio.2005282.g001.png){fig-alt="2 figures showing point estimates and confidence intervals for estimates of proportion of published papers with pseudoreplication issues"}](https://doi.org/10.1371/journal.pbio.2005282.g001)

::: aside
From: Lazic et al. (2018) *What exactly is ‘N’in cell culture and animal
experiments?.* PLoS Biology.
https://doi.org/10.1371/journal.pbio.2005282
:::

## Estimation - Summary

-   Parameter estimates are influenced by chance.

-   The sampling distribution considers what could have happened upon
    sampling.

-   The standard error is the standard deviation of the sampling
    distribution.

-   If producing a real sampling distribution is not possible, a
    bootstrapping distribution based on the original sample tends to be
    very similar to the real sampling distribution.

-   Confidence intervals bound plausible parameter values.

-   Uncertainty decreases and precision increases with sample size.

-   Add error bars to graphs to illustrate standard errors or confidence
    intervals. Make sure to clarify which is being illustrated in the
    figure legend.

## That's all for today

![From:
makeameme.org](images/forrestgump.png){fig-alt="\"Forrest says \"And that's all I wanted to say about that\""
width="347"}
